A Simple Type Theory for Category Theory

### 2022-02-11 paper/lipics-v2021-sample-article.tex
intro update

Experienced category theorists have no trouble with this, as they are
familiar with a great many principles for constructing transformations
that are guaranteed to be natural. Consider an analogy with
calculus. When a mathematician writes a formula such as

$$\sin(f(x) - g(x))$$

They do not need to spend time with an explicit $\epsilon,\delta$
proof that the function is continuous or furthermore smooth. Instead,
they follow syntactic principles that compositions of continuous
functions are continuous.

There are similar syntactic notions for guaranteeing that
transformations are natural.

\subsection{Profunctors}

@@

@@

>>>>

More experienced users of category theory in programming language
theory find that for modeling computational effects such as general
recursion or indexed-state, or modeling logics for parametricity,
gradual typing, or step-indexed logical relations involves not just
ordinary category theory, but more general notions such as enriched,
internal or multi-categories. However, model construction theorems and
especially soundness and completeness theorems in these areas will
require their own, even more complex versions of the same common
lemmas of category theory, in addition to the domain-specific results
such as existence of fixed points.

At this level of generality, the proofs can be difficult to
understand, as they often involve some combination of the original
category theoretic proof with some additional side-conditions that
everythinng is in fact appropriately step-indexed, continuous etc.
%
The solution to taming this complexity is to prove these general
theorems at a higher level of generality, for instance, proving them
in some abstract 2-category of enriched/internal categories, functors
and transformations.
%
For proofs in generalized category theory specifically, the notion of
a \emph{virtual equipment} has been shown to be occupy an algebraic
sweet-spot of being complex enough to express most basic constructions
(functors, profunctors, natural transformations, adjunctions, kan
extensions, etc) but also general enough to encompass internal,
enriched and ``generalized multi-categories'', which include ordinary
flavors of multi-categories like the cartesian multi-categories that
model sequents in natural deduction, but even structures such as
topological spaces.
%
However, there are two difficulties with this approach.
%
First, the abstract arguments can be difficult to understand, and can
seem quite hard to connect to the original argument for ordinary
category theory.
%
Second, the algebraic structure of a virtual equipment is quite
unwieldy to manipulate. The equivalent of a 2-cell is a kind of
``multi-2-cell'' with many inputs and a single output, in
diagrammatic form these are rendered as

\[ % https://tikzcd.yichuanshen.de/#N4Igdg9gJgpgziAXAbVABwnAlgFyxMJZABgBpiBdUkANwEMAbAVxiRAGEB9YkAX1PSZc+QigCM5KrUYs2XMXwEgM2PASIAmSdXrNWiDpw2LBqkUQDM26XrYAdOwygQcCfqeHqUAFmu7ZBlzAYADUYrwmykJqoiSkYlL++iCRKp6xvgk6Msl8UjBQAObwRKAAZgBOEAC2SBIgOBBIWjYBIABKnAruIJU1zdSNSFatyZ3GPX21iCNDiL6jbJ2Ek1XTZA1NiPVJS9yRU0gbcwCs2bYGAGIHa0hnm0gAbOdtAMo3-fODW8+LBgDieV4QA
\begin{tikzcd}
C_0 \arrow[r, "R_0"] \arrow[d, "F"] & C_1 \arrow[r, "R_1"] & C_2 \arrow[r, "R_2"] & \ldots \arrow[r, "R_n"] & C_{n+1} \arrow[d, "G"] \\
{} \arrow[rrrr, "S"]                &                      &                      &                         & {}                    
\end{tikzcd} \]

This 2-cell has one output pro-arrow, many input pro-arrows which must
match the source from one to the target object of the next and a left
and right ``frame'' arrow. These can be composed in a manner similar
to multi-categories, where you can compose this with many input
2-cells, one for each input pro-arrow.
%
These then have complex associativity and unit rules and side
conditions that say the framing arrows compose alongside the 2-cells.

We solve both of these issues by developing an alternative \emph{type
theoretic syntax} for the constructions in a virtual equipment.
%
In this syntax the ``multi-2-cell'' above instead becomes a natural
transformation judgment
\[ \Phi \vdash t : R[F(x);G(y)] \]
We write natural transformations in a natural deduction style: a
transformation is written as a ``parameterized'' element of the output
profunctor. Furthermore the restriction operators for vertical arrows
into horizontal arrows is naturally modeled by a syntactic
substitution.
%
This interpretation of restriction as substitution both matches the
semantic intuition for the restriction operation as well as greatly
simplifies reasoning about composition of transformations.

%% Experienced category theorists have no trouble with this, as they are
%% familiar with a great many principles for constructing transformations
%% that are guaranteed to be natural. Consider an analogy with
%% calculus. When a mathematician writes a formula such as

%% $$\sin(f(x) - g(x))$$

%% They do not need to spend time with an explicit $\epsilon,\delta$
%% proof that the function is continuous or furthermore smooth. Instead,
%% they follow syntactic principles that compositions of continuous
%% functions are continuous.

%% There are similar syntactic notions for guaranteeing that
%% transformations are natural.

\subsection{Category Theory with Profunctors}

@@

\subsection{Paper Overview}

To demonstrate the utility of SCT, we proceed as follows
\begin{enumerate}
\item In \cref{sec:syntax} we introduce the syntax of Simple Category
  Theory.
\item In \cref{sec:synthetic} we show how common theorems of category
  theory such as the (co)-Yoneda lemma, equivalence of different
  formulations of adjunctions, and preservation of limits can all be
  cleanly reproduced in SCT.
\item In \cref{sec:internal}, we show how we can use SCT to define
  \emph{internal categories}, which allows us to prove additional
  theorems that are not formalizable as synthetic theorems.
\item In \cref{sec:semantics}, we show that the syntax of SCT is sound
  for models in an arbitrary virtual double category with
  corresponding universal constructions.
\item In \cref{sec:discussion}, we discuss related type theories and
  potential extensions to model multi-variable and higher-order
  categorical constructions.
\end{enumerate}

@@

\begin{definition}[Psuedo Double Category]

\end{definition}

\begin{definition}[Restriction]

\end{definition}

\begin{definition}[Unit]

\end{definition}
